---
title: 'Data Mining Homework 2: Premier League'
author: "Max Buckley"
date: "22/03/2016"
output: pdf_document
---

In this assignment I aim to explore throughly the provided dataset of Premier League soccer matches. From the brief the objective is:

>Your task is to analyse Premier League match data provided since 2010 up to and including last weekend (6 March 2016). You should attempt to uncover interesting and potentially useful patterns and rules which might be used to determine the outcome of the remaining matches in the 2016 league.

I will break my analysis into its logical steps so that I start with data cleaning and preperation, progress to data exploration, and then lastly attempt data modelling and prediction. I will include all my code in the interests of reproducable research.

An interesting thing to note is I have absolutely no interest in soccer I may be unaware of or ask naive questions of the data vis-a-vis the soccer afficianados in the class.

#Data Cleaning

```{r}
# Dependancies.
library("plyr")

# Clear R environment.
rm(list=ls())
# Set working directory .
setwd("~/UCD/DataMining/HW2/")

output_dir <- "CleanedData"
input_dir <- "InputData"

input_csv_files <- dir(input_dir)

# List to store all the data frames.
df_list <- list()
for(i in 1:length(input_csv_files)){
  file <- input_csv_files[i]
  df <- read.csv(paste0(input_dir, "/", file),
                 stringsAsFactors=FALSE)
  # Store a season indicator.
  df$Season <- gsub(".csv", "", file)
  df_list[[i]] <- df
}

# Combine all the data frames into one.
cleaned_data <- rbind.fill(df_list[1:i])
# Format date as proper R date object.
cleaned_data$Date <- as.Date(cleaned_data$Date, "%d/%m/%y")
# Factor desired character columns.
cleaned_data$FTR <- as.factor(cleaned_data$FTR)
cleaned_data$HTR <- as.factor(cleaned_data$HTR)
cleaned_data$HomeTeam <- as.factor(cleaned_data$HomeTeam)
cleaned_data$AwayTeam <- as.factor(cleaned_data$AwayTeam)
cleaned_data$Referee <- as.factor(cleaned_data$Referee)
cleaned_data$Season <- as.factor(cleaned_data$Season)

# Remove Division as it is a constant. It adds no information.
cleaned_data$Div <- NULL

#Calculate points
HomePoints<-rep(1, nrow(cleaned_data))
HomePoints[cleaned_data$FTR=="H"] <- 3
HomePoints[cleaned_data$FTR=="A"] <- 0
cleaned_data$HomePoints <- HomePoints
AwayPoints<-rep(1,nrow(cleaned_data))
AwayPoints[cleaned_data$FTR=="H"] <- 0
AwayPoints[cleaned_data$FTR=="A"] <- 3
cleaned_data$AwayPoints <-AwayPoints
cleaned_data$TotalGoals <-cleaned_data$FTHG + cleaned_data$FTAG


# Save as output to use later. Use .Rda format to preserve R
# type information in addition to data values.
save(cleaned_data, file = paste0(output_dir, "/",
                                 "cleaned_data.Rda"))
```
With some of my data preprocessing I now had a better idea of what my data source looks like and how it is strucutred.

Naively one would assume that the match result and teams playing would appear to be the most imortant factors. So a question to ask is which teams are the best historically. How are these different teams doing?

Given the dataset is currently in wide form and quite difficult to work with I decided to switch it to long form. So rather than having the winner and loser in one row I would have one row for each side and I could then more easily aggregate it.

```{r message=FALSE, warning=FALSE}
# Libraries.
# Best plotting library ever
library("ggplot2")
# For data munging
library("reshape")
# More data munging
library("reshape2")
# Same
library("plyr")
# Recursive Partitioning decision trees
library("rpart")
# The party package provides nonparametric regression trees
# for nominal, ordinal, numeric, censored, and multivariate
# responses.
library("party")
# Apriori algorithm
library("arules")

# Clear R environment.
rm(list=ls())

# Set working directory .
setwd("~/UCD/DataMining/HW2/")
# This creates an object called cleaned_data in our environment.
load("CleanedData/cleaned_data.Rda")

#Take a subset of our data
subset <- cleaned_data[,c(1:6,10,71,75,76,77)]
sub_data <- melt(subset, id=c("Date", "FTHG", "FTAG", "FTR",
                              "Referee", "Season", "HomePoints",
                              "AwayPoints", "TotalGoals"))
sub_data <- rename(sub_data, c("value"="TeamName"))
sub_data <- rename(sub_data, c("variable"="HomeOrAway"))

# Calculate the respective points for each team.
sub_data$Points <- 0
sub_data$Points[sub_data$HomeOrAway=="AwayTeam"] <- sub_data$AwayPoints[
  sub_data$HomeOrAway=="AwayTeam"]
sub_data$Points[sub_data$HomeOrAway=="HomeTeam"] <- sub_data$HomePoints[
  sub_data$HomeOrAway=="HomeTeam"]

sub_data$Goals <-0
sub_data$Goals[sub_data$HomeOrAway=="AwayTeam"] <- sub_data$FTAG[
  sub_data$HomeOrAway=="AwayTeam"]
sub_data$Goals[sub_data$HomeOrAway=="HomeTeam"] <- sub_data$FTHG[
  sub_data$HomeOrAway=="HomeTeam"]

opposition <-cbind.data.frame(rep(cleaned_data$AwayTeam, 2),
                              rep(cleaned_data$HomeTeam, 2))
vec<-vector()
for(i in 1:nrow(opposition)){
  if(sub_data$TeamName[i]==opposition[i, 1]){
   vec[i] <- as.character(opposition[i, 2])
  }
  else{
    vec[i] <- as.character(opposition[i, 1])
  }
}
sub_data$Opponent <-as.factor(vec)

# Introduce a calculate metric. goal ratio. Which tells us
# what proportion of the goals in that match that team scored.
# To prevent division by zero NaNs. The round later removes this.
ratio_offset<-1e-6
sub_data$GoalRatio<-round((sub_data$Goals + ratio_offset)/
                          (sub_data$TotalGoals + ratio_offset),
                          digits=2)

head(sub_data)
```

# Data Exploration
```{r message=FALSE, warning=FALSE}
#Take a subset again
interest_cols<-c("Season", "Date", "HomeOrAway", "Referee",
                 "TeamName", "Opponent", "Points", "Goals",
                 "TotalGoals", "GoalRatio")

formed_data <- sub_data[, interest_cols]

team_points <-aggregate(sub_data$Points, by=list(
  sub_data$TeamName, sub_data$Opponent), FUN=sum)

team_points_mat <-cast(team_points, Group.1~Group.2, sum)

# Need to calculate what fraction of the points the team got of what they
# could have gotten in total.
team_point_ratios <-round(
  (team_points_mat[,-1])/(
    (table(cleaned_data$HomeTeam,cleaned_data$AwayTeam) +
       table(cleaned_data$AwayTeam,cleaned_data$HomeTeam))*3),4)

#Update row names
row.names(team_point_ratios)<-colnames(team_point_ratios)

# Replace all 0s.
team_point_ratios <- replace(team_point_ratios, is.na(team_point_ratios), 0)

# Heat map clusters the matrix heirarchically and colours the cells by the value intensity.
heatmap(as.matrix(team_point_ratios))

# Melt the team point raios to get a historical point record.
# A variable that tells use how well this team has done
# against this opponent.
historical_point_record<-melt(cbind(row.names(
  team_point_ratios), team_point_ratios))
colnames(historical_point_record)<-c("TeamName", "Opponent",
                                     "PointRatio")
```

From this heatamap we can see some instant information .... PLEASE FILL OUT.



```{r}
numerical <- formed_data[,-c(1,2,3,4,5,6)]
sum_mat <- aggregate(numerical, by=list(formed_data$Season, formed_data$TeamName),
                     FUN=sum)
mean_mat <- aggregate(numerical, by=list(formed_data$Season, formed_data$TeamName),
                      FUN=mean)
sum_mat$GoalRatio<-mean_mat$GoalRatio
sum_mat <- rename(sum_mat, c("Group.1"="Season","Group.2"="TeamName"))
sum_mat <-sum_mat[order(as.character(sum_mat$Season),
                        sum_mat$Points,
                        decreasing=TRUE),]
row.names(sum_mat) <- 1:120
sum_mat<-ddply(sum_mat, .(Season), transform, Rank = rank(-Points,
                                                          ties.method="first"))

# Lets take a look at total points by team per season. 
p <- ggplot(sum_mat, aes(Season, Points, group=TeamName,
                        colour=TeamName))
p + geom_path(alpha=1)+ labs(title = "Team Points by Season") +
  guides(col = guide_legend(nrow = 16)) +
  theme(axis.text.x=element_text(size=8))
```

Of course we immeadiatly see multiple problems with this chart. One is that the 2015/16 season 
is incomplete so all the lines collapse to the center. Also maybe it is a better question to
ask what is the rank of any given team in a particular year.

TBC: